Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2021-10-27 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

\usetheme[progressbar=frametitle]{metropolis}
\usepackage{appendixnumberbeamer}
\usepackage{multicol}
\usepackage{booktabs}
\usepackage[scale=2]{ccicons}
\usepackage[style=authoryear]{biblatex}
\bibliography{demo.bib}
\usepackage{pgfplots}
\usepackage{bbm}
\usepackage{import}
\usepackage{graphics}

@@

\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}

@@

\subtitle{Hanna Glenn}
 \date{\today}
% \date{}
%\author{Presented by: Hanna Glenn}
% \institute{Center for modern beamer themes}
% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}

@@

\begin{frame}{Table of Contents}
  \setbeamertemplate{section in toc}[sections numbered]
  \tableofcontents%[hideallsubsections]
\end{frame}

@@

\begin{frame}{What is an EHR?}
\begin{itemize}
    \item Technology purchased and implemented by hospitals
\end{itemize}

\begin{itemize}
    \item Used for:
    \vspace{3mm}
    \begin{itemize}
        \item Electronic patient records
        \vspace{3mm}
        \item More complex (and expensive) EHRs allow for decision making assistance
        \vspace{3mm}
        \item Continuing to evolve over time
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{HIT: Great (Expected) Potential in Healthcare}
\begin{alertblock}{Cost Saving}
\begin{itemize}
    \item Possible cost reduction of hundreds of billions of dollars \\ (\cite{hillestad2005})
\end{itemize}
\end{alertblock}

\begin{alertblock}{Quality Improvement}
\begin{itemize}
    \item Improved efficiency, patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
    \item Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs \nocite{hitech}
\end{itemize}
\end{alertblock}

\textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} (\cite{stats})

@@

\begin{frame}[fragile]{What does this mean for physicians?}
Immediate Effects:
\begin{itemize}
    \item Large startup costs (time, productivity)
    \item Possible loss of autonomy
\end{itemize}

Long Run Effects:
\begin{itemize}
    \item Productivity could increase or decrease
    \item Progressing loss of autonomy as technology progresses
\end{itemize}

@@

\begin{frame}[noframenumbering]{Physician Burnout in the Media}
\begin{center}
    \includegraphics[scale=.4]{graphics/News Clip1.PNG}
\end{center}
\end{frame}

\begin{frame}[noframenumbering]{Physician Burnout in the Media}
\begin{center}
    \includegraphics[scale=.5]{graphics/News Clip2.PNG}
\end{center}
\end{frame}

\begin{frame}[noframenumbering]{Physician Burnout in the Media}
\begin{center}
    \includegraphics[scale=.45]{graphics/News Clip3.PNG}
\end{center}
\end{frame}

@@

\begin{frame}{Literature}

@@

\begin{frame}{What I Want to Capture}
\begin{itemize}
    \item Initial Implementation
    \begin{itemize}
        \item Productivity gains/losses
        \item Immediate changes to labor market decisions: retirement or place of work
    \end{itemize}
    \item Loss of autonomy and cost of learning/using new technology
    \begin{itemize}
        \item Could cause physicians to move to settings where they have more control
    \end{itemize}
\end{itemize}

\begin{frame}{About the Data}
    Main measurements come from \underline{CMS Shared Patient Data} (2009-2015)
    \begin{itemize}
        \item Lists 2 NPIs and the number of billings those two entities shared in the same day
    \end{itemize}

@@

     Total hospital patients: 
    \begin{itemize}
        \item Sum of shared patients with all hospitals
    \end{itemize}

@@

     Indicator for switching work setting:
    \begin{itemize}
        \item Positive billings shared with all entities but shared entities with hospitals becomes 0
        \item \color{gray}{SK$\&$A database of office-based physicians}
    \end{itemize}

@@

Physician-level treatment variable that captures exposure to \textit{any} EHR
\begin{itemize}
    \item Individual hospital EHR use comes from AHA survey
    \item Treatment turns on the first time any of the shared hospitals implements an EHR
\end{itemize}
\end{frame}

@@

Medical school graduation date, gender: \underline{Physician Compare}

@@

\begin{frame}{Summary Statistics}
\centering
    \includegraphics[scale=.8]{Objects/sumstats.pdf}
\end{frame}

@@

>>>>

\usetheme[progressbar=frametitle]{metropolis}
\usepackage{appendixnumberbeamer}
\usepackage{pgfpages}
\usepackage{booktabs}
\usepackage{xcolor}

\usepackage{pgfplots}

@@

\setbeameroption{show notes on second screen=right}
\setbeamertemplate{note page}{\pagecolor{yellow!5}\insertnote}\usepackage{palatino}
\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}

@@

\subtitle{Hanna Glenn}
 \date{\today}

\setbeamertemplate{itemize item}{\scalebox{.6}{$\blacktriangleright$ }}     
\setbeamertemplate{itemize subitem}{\scalebox{.8}{$\centerdot$ }}

@@

\begin{frame}{Table of Contents}
  \setbeamertemplate{section in toc}[sections numbered]
  \tableofcontents%[hideallsubsections]

  \note[item]{Roadmap of the presentation}
\end{frame}

@@

\begin{frame}[fragile]{What is an EHR?}
\begin{itemize}
    \item Technology purchased and implemented by hospitals
\end{itemize}

\begin{itemize}
    \item Used for:
    \vspace{3mm}
    \begin{itemize}
        \item Electronic patient records
        \vspace{3mm}
        \item More complex (and expensive) EHRs allow for decision making assistance
        \vspace{3mm}
        \item Continuing to evolve over time
    \end{itemize}
\end{itemize}

\note[item]{Get everyone on the same page about how to think about EHRs}
\note[item]{A product in the form of technology that is sold to hospitals (and others, but we focus on hospital setting)}
\note[item]{The most standard element of an EHR is that it converts paper records to electronic records}
\note[item]{New elements of EHRs: interropability, patient interface, decision-making assistance}

@@

\begin{frame}[fragile]{HIT: Great (Expected) Potential in Healthcare}
\begin{alertblock}{Cost Saving}
\begin{itemize}
    \item Possible cost reduction of hundreds of billions of dollars \\ \scriptsize (Hillestad et al 2005, HA)
\end{itemize}
\end{alertblock}

\begin{alertblock}{Quality Improvement}
\begin{itemize}
    \item Improved efficiency, patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
    \item Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs 
\end{itemize}
\end{alertblock}

\textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} \scriptsize (Health IT Dashboard)

\note[item]{Why do we care about EHRs or health information technology in the first place?}

@@

\begin{frame}[fragile]{What could this mean for physicians?}
Physicians are the ones who put this technology to use, how they respond will partially determine the technology's impacts

\vspace{3mm}

Immediate Effects:
\begin{itemize}
    \item Large startup costs (time, productivity)
    \item Possible loss of autonomy
\end{itemize}

Long Run Effects:
\begin{itemize}
    \item Productivity could increase
    \item Progressing loss of autonomy as technology progresses
\end{itemize}

\begin{frame}[noframenumbering]{Physician Response to EHRs}
\begin{center}
    \includegraphics[scale=.4]{graphics/News Clip1.PNG}
\end{center}
\end{frame}

\begin{frame}[noframenumbering]{Physician Response to EHRs}
\begin{center}
    \includegraphics[scale=.5]{graphics/News Clip2.PNG}
\end{center}
\end{frame}

\begin{frame}[noframenumbering]{Physician Response to EHRs}
\begin{center}
    \includegraphics[scale=.35]{graphics/News Clip3.PNG}
\end{center}
\end{frame}

@@

\begin{frame}{Literature}
\footnotesize 
\begin{columns}
    \column{0.34\textwidth}
        \centering
        \underline{ EHR $\rightarrow$ Healthcare }
        \begin{itemize}
            \item Small improvements for severe health conditions \\ \vspace{1mm}
            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)

            \footnotesize

            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
            \tiny (Appari et al 2013)

            \footnotesize

            \item No decrease in cost \\ \vspace{1mm}
            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)

            \footnotesize

            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)

        \end{itemize}

    \column{0.33\textwidth}
    \color{gray}
        \centering
        \underline{ Technology $\rightarrow$ Labor }
        \begin{itemize}
        \color{gray}
            \item Erosion effect vs. wage effect \\ \vspace{1mm}
            \tiny (Zeira $\&$ Joseph 2011)

            \footnotesize

            \item Cause retirement? \\ \vspace{1mm}
            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)

            \footnotesize

            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
            \tiny (Willis and Hudomiet 2020)
        \end{itemize}

    \column{0.33\textwidth}
        \color{gray}
        \centering
        Physician Labor \underline{ Determinants }
        \begin{itemize}
        \color{gray}
            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
            \tiny (Baltagi et al 2005, Sloan 2018)

            \footnotesize

            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
            \tiny (Bahrami 2002)

            \footnotesize

            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
            \tiny (Dewa et al 2014)
        \end{itemize}

\end{columns}
\end{frame}

\begin{frame}[noframenumbering]{Literature}
\footnotesize 
\begin{columns}
    \column{0.34\textwidth}
    \color{gray}
        \centering
        \underline{ EHR $\rightarrow$ Healthcare }
        \begin{itemize}
        \color{gray}
            \item Small improvements for severe health conditions \\ \vspace{1mm}
            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)

            \footnotesize

            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
            \tiny (Appari et al 2013)

            \footnotesize

            \item No decrease in cost \\ \vspace{1mm}
            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)

            \footnotesize

            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)

        \end{itemize}

    \column{0.33\textwidth}
        \centering
        \underline{ Technology $\rightarrow$ Labor }
        \begin{itemize}
            \item Erosion effect vs. wage effect \\ \vspace{1mm}
            \tiny (Zeira $\&$ Joseph 2011)

            \footnotesize

            \item Cause retirement? \\ \vspace{1mm}
            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)

            \footnotesize

            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
            \tiny (Willis and Hudomiet 2020)
        \end{itemize}

    \column{0.33\textwidth}
    \color{gray}
        \centering
        Physician Labor \underline{ Determinants }
        \begin{itemize}
        \color{gray}
            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
            \tiny (Baltagi et al 2005, Sloan 2018)

            \footnotesize

            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
            \tiny (Bahrami 2002)

            \footnotesize

            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
            \tiny (Dewa et al 2014)
        \end{itemize}
\end{columns}
\end{frame}

\begin{frame}[noframenumbering]{Literature}
\footnotesize 
\begin{columns}
    \column{0.34\textwidth}
    \color{gray}
        \centering
        \underline{ EHR $\rightarrow$ Healthcare }
        \begin{itemize}
        \color{gray}
            \item Small improvements for severe health conditions \\ \vspace{1mm}
            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)

            \footnotesize

            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
            \tiny (Appari et al 2013)

            \footnotesize

            \item No decrease in cost \\ \vspace{1mm}
            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)

            \footnotesize

            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)

        \end{itemize}

    \column{0.33\textwidth}
    \color{gray}
        \centering
        \underline{ Technology $\rightarrow$ Labor }
        \begin{itemize}
        \color{gray}
            \item Erosion effect vs. wage effect \\ \vspace{1mm}
            \tiny (Zeira $\&$ Joseph 2011)

            \footnotesize

            \item Cause retirement? \\ \vspace{1mm}
            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)

            \footnotesize

            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
            \tiny (Willis and Hudomiet 2020)
        \end{itemize}

    \column{0.33\textwidth}
        \centering
        Physician Labor \underline{ Determinants }
        \begin{itemize}
            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
            \tiny (Baltagi et al 2005, Sloan 2018)

            \footnotesize

            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
            \tiny (Bahrami 2002)

            \footnotesize

            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
            \tiny (Dewa et al 2014)
        \end{itemize}
\end{columns}
\end{frame}

\begin{frame}{What I Want to Capture}
\begin{itemize}
    \item Initial Implementation
    \vspace{3mm}
    \begin{itemize}
        \item Productivity gains/losses?
        \vspace{3mm}
        \item Retirement or change in place of work?
        \vspace{3mm}
    \end{itemize}
    \item Do the effects change over time after the technology is implemented?

\end{itemize}

@@

\begin{frame}{About the Data}
    Main measurements come from \underline{CMS Shared Patient Data} (2009-2015)
    \begin{itemize}
        \item Lists 2 NPIs and the number of billings those two entities shared
    \end{itemize}
    \vspace{3mm}

@@

     Hospital patients: 
    \begin{itemize}
        \item Sum of shared patients with all hospitals
    \end{itemize}

@@

     Indicator for switching work setting:
    \begin{itemize}
        \item Positive billings shared with all entities but shared entities with hospitals becomes 0
        \item Database of office-based physicians
    \end{itemize}

@@

Physician-level treatment variable that captures exposure to \textit{any} EHR
\vspace{3mm}
\begin{itemize}
    \item Individual hospital EHR use comes from AHA survey
    \vspace{3mm}
    \item Treatment turns on the first time any of the shared hospitals implements an EHR
\end{itemize}
\end{frame}

@@

Medical school graduation date: \underline{Physician Compare}

@@

\begin{frame}{Summary Statistics}
\centering
    \includegraphics[scale=.8]{Objects/sumstats.pdf}

    \note[item]{Need to divide n by 7 to get actual number of physicians}
\end{frame}

@@

\section{Continuing Work}

\begin{frame}{Plans}
    \begin{itemize}
        \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicaid data
    \end{itemize}
\end{frame}
